Elaboration with First-Class Implicit Function Types

### 2020-05-11 paper.tex
revs

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. In this section, we use it to build the backbone of our
algorithm and illustrate the key issues. We extend this elaborator in Section
\ref{sec:extending_elaboration}.
\begin{figure}[h]
\begin{alignat*}{4}
  t,\,u,\,v,\,A,\,B,\,C\, :&:=\quad  && x\hspace{8em}              & \text{variable}                 &  \\
                           & |       && (x : A)\to B               & \text{function type}            &  \\
                           & |       && \{x : A\}\to B             & \text{implicit function type}   &  \\
                           & |       && t\,u                       & \text{application}              &  \\
                           & |       && t\,\{u\}                   & \text{implicit application}     &  \\
                           & |       && \lambda\,x.\, t            & \text{lambda abstraction}       &  \\
                           & |       && \lambda\,\{x\}.\,t         & \text{implicit abstraction}     &  \\
                           & |       && \U                         & \text{universe}                 &  \\
                           & |       && \slet\,x : A = t\,\sin\, u & \text{let-definition}           &  \\
                           & |       && \_                         & \text{hole for inferred term}   &
\end{alignat*}
\caption{Syntax of the surface language.}
\label{fig:surface}
\end{figure}

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have implicit
argument insertion as a helper function, which is defined by recursion on core
types, and which is used as a post-processing step after inference. First, about
the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insert]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
\end{mathpar}
\begin{alignat*}{3}
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail
\end{alignat*}
\pagebreak
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

@@

We revisit now the $polyList$ example from Section \ref{sec:introduction}. We
assume the following:
\[
List : \U \to \U \qquad nil : \{A : \U\} \to List\,A \qquad
cons : \{A : \U\} \to A \to List\,A \to List\,A
\]
In the following, we present a trace of checking $cons\,(\lambda\,x.\,x)\,nil$
at type $List\,(\{A : \U\}\to A \to A)$. We omit context and metacontext
parameters everywhere, and mark recursive calls by indentation. We also omit
some checking, inference, implicit insertion and unification calls which are not
essential for illustration.
\begin{alignat*}{3}
  & \scriptstyle{0 }\qquad\qquad && \echeckt{cons\,(\lambda\,x.\,x)\,nil}{(List\,(\{A : \U\}\to A \to A))}
     \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil} \\
  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)}\\
  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}\\
  & \scriptstyle{4 }  && \qquad\quad = cons\,\{\alpha_0\}\,:\,\alpha_0 \to List\,\alpha_0 \to List\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = cons\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x) : List\,(\alpha_1\to\alpha_1) \to List\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{8 }  && \qquad \echeckt{nil}{(List\,(\alpha_1\to\alpha_1))}\\
  & \scriptstyle{9 }  && \qquad = nil\,\{\alpha_1\to\alpha_1\}\\
  & \scriptstyle{10} && \quad = cons\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x)\,(nil\,\{\alpha_1\to\alpha_1\}): List\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{11} && \quad \unify\,(List\,(\{A : \U\}\to A \to A))\,(List\,(\alpha_1\to\alpha_1))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{13} && \qquad = \fail
\end{alignat*}

@@

>>>>

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. Our main goal with the elaborator is to have a setup
which is as simple as possible, but sufficient to illustrate the enhancements in
Section \ref{sec:extending_elaboration}, and make it clear that this could be
used to conservatively extend existing dependently typed languages. Hence, we do
not have more fine-grained control over metavariable insertion, nor the advanced
unification algorithms used in Agda \cite{abel2011higher}; all of these are
orthogonal improvements of the presented system.

\begin{figure}[h]
\begin{alignat*}{4}
  t,\,u,\,v,\,A,\,B,\,C\, :&:=\quad  && x\hspace{8em}              & \text{variable}                 &  \\
                           & |       && (x : A)\to B               & \text{function type}            &  \\
                           & |       && \{x : A\}\to B             & \text{implicit function type}   &  \\
                           & |       && t\,u                       & \text{application}              &  \\
                           & |       && t\,\{u\}                   & \text{implicit application}     &  \\
                           & |       && \lambda\,x.\, t            & \text{lambda abstraction}       &  \\
                           & |       && \lambda\,\{x\}.\,t         & \text{implicit abstraction}     &  \\
                           & |       && \U                         & \text{universe}                 &  \\
                           & |       && \slet\,x : A = t\,\sin\, u & \text{let-definition}           &  \\
                           & |       && \_                         & \text{hole for inferred term}   &
\end{alignat*}
\caption{Syntax of the surface language.}
\label{fig:surface}
\end{figure}

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have implicit
argument insertion as a helper function, which is defined by recursion on core
types, and which is used as a post-processing step after inference. First, about
the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insert]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
\end{mathpar}
\begin{alignat*}{3}
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail
\end{alignat*}
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

@@

We revisit now the $polyList$ example from Section \ref{sec:introduction}. We
assume the following:
\[
List : \U \to \U \qquad nil : \{A : \U\} \to List\,A \qquad
cons : \{A : \U\} \to A \to List\,A \to List\,A
\]
In Figure \ref{fig:polylist}, we present a trace of checking $cons\,(\lambda\,x.\,x)\,nil$
at type $List\,(\{A : \U\}\to A \to A)$. We omit context and metacontext
parameters everywhere, and mark recursive calls by indentation. We also omit
some checking, inference, implicit insertion and unification calls which are not
essential for illustration.

\begin{figure}
\begin{alignat*}{3}
  & \scriptstyle{0 }\qquad\qquad && \echeckt{cons\,(\lambda\,x.\,x)\,nil}{(List\,(\{A : \U\}\to A \to A))}
     \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil} \\
  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)}\\
  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}\\
  & \scriptstyle{4 }  && \qquad\quad = cons\,\{\alpha_0\}\,:\,\alpha_0 \to List\,\alpha_0 \to List\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = cons\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x) : List\,(\alpha_1\to\alpha_1) \to List\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{8 }  && \qquad \echeckt{nil}{(List\,(\alpha_1\to\alpha_1))}\\
  & \scriptstyle{9 }  && \qquad = nil\,\{\alpha_1\to\alpha_1\}\\
  & \scriptstyle{10} && \quad = cons\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x)\,(nil\,\{\alpha_1\to\alpha_1\}): List\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{11} && \quad \unify\,(List\,(\{A : \U\}\to A \to A))\,(List\,(\alpha_1\to\alpha_1))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{13} && \qquad = \fail
\end{alignat*}
\caption{$\mi{polyList}$ elaboration.}
\label{fig:polylist}
\end{figure}

@@

\end{figure}
